---
title: "NYPD Shooting Incident Project"
author: "C. Mill"
date: "2026-02-12"
output: html_document
---

# Step 1: Import the data

This project uses the NYPD Shooting Incident Data (Historic).  
It includes shooting date/time, borough, and whether the incident was marked as a statistical murder.


`ARCHIVED_NYPD_Shooting_Incident_Data__Historic_.csv`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(lubridate)
library(janitor)
library(yardstick)
library(scales)
```

```{r import}
shoot_raw <- read_csv(
  "ARCHIVED_NYPD_Shooting_Incident_Data__Historic_.csv",
  show_col_types = FALSE
) %>%
  clean_names()

glimpse(shoot_raw)
```

# Step 2: Tidy the data

We convert the date column into a real Date type and create a month column for plotting.

```{r tidy}
shoot <- shoot_raw %>%
  mutate(
    occur_date = mdy(occur_date),
    month = floor_date(occur_date, "month"),
    boro = as.factor(boro),
    statistical_murder_flag = as.integer(statistical_murder_flag)  # TRUE/FALSE -> 1/0
  )

summary(shoot)
```

## Missing data check

```{r missing}
missing_tbl <- shoot %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "missing_values") %>%
  arrange(desc(missing_values))

missing_tbl
```

If a plot needs a variable (like borough), I will filter out missing values only for that plot.  
For the model, I will keep only rows that have the variables needed.

# Step 3: Visualizations + short analysis

## Plot 1: Shootings over time (monthly)

```{r plot1}
monthly <- shoot %>%
  filter(!is.na(month)) %>%
  count(month)

ggplot(monthly, aes(month, n)) +
  geom_line() +
  labs(
    title = "Monthly Shooting Incidents",
    x = "Month",
    y = "Number of incidents"
  ) +
  scale_y_continuous(labels = comma)
```

**Quick thought:** If the line goes up or down, it suggests shootings changed over time.  
A question to investigate later: do trends differ by borough?

## Plot 2: Shootings by borough

```{r plot2}
boro_counts <- shoot %>%
  filter(!is.na(boro)) %>%
  count(boro, sort = TRUE)

ggplot(boro_counts, aes(reorder(boro, n), n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Shooting Incidents by Borough",
    x = "Borough",
    y = "Number of incidents"
  ) +
  scale_y_continuous(labels = comma)
```

**Quick thought:** Borough totals may differ due to population size, environment, and reporting/policing differences.

# Step 3 (required): One simple model


```{r model}
model_df <- shoot %>%
  filter(!is.na(statistical_murder_flag), !is.na(boro))

set.seed(123)
idx <- sample.int(nrow(model_df), size = floor(0.8 * nrow(model_df)))
train <- model_df[idx, ]
test  <- model_df[-idx, ]

m1 <- glm(
  statistical_murder_flag ~ boro,
  data = train,
  family = binomial()
)

summary(m1)
```

## Model evaluation

```{r eval}
test_prob <- predict(m1, newdata = test, type = "response")
test_pred <- if_else(test_prob > 0.5, 1, 0)

results <- tibble(
  truth = factor(test$statistical_murder_flag, levels = c(0, 1)),
  estimate = factor(test_pred, levels = c(0, 1)),
  prob = test_prob
)

conf_mat(results, truth, estimate)
roc_auc(results, truth, prob)
```

# Step 4: Conclusion + Bias

## Conclusion
This project imported and cleaned the NYPD Shooting Incident dataset and created two visualizations.
The plots show how shootings vary over time and differ by borough.
A simple logistic regression model used borough to predict whether an incident was marked as a statistical murder.
This model shows association only and does not prove cause.

## Possible bias in the data
- The dataset includes only incidents recorded by NYPD (unreported incidents are missing).
- Some fields may be missing or recorded inconsistently.
- “Statistical murder” is an administrative label and may reflect classification practices.

## Personal bias and how I reduced it
My bias could be assuming patterns reflect true behavior rather than reporting/system effects.
